A North Sámi to South Sámi machine translation prototype

### 2014-02-09 paper/paper.tex
e
fonts
add abstract
updates
Meir tekst.

\usepackage[small,bf]{caption}
\usepackage{xltxtra}
\usepackage{fontspec}

@@

\setromanfont{Times New Roman}
%\setmonofont{FreeMono}

@@

@@

%pivot translation references --babych paper

@@

Norwegian North Germanic. The Saami languages constitute the
westernmost branch of the Uralic language family. They possess most of
the classical Uralic characteristica: A rich verbal morphology (3
numbers and persons, a rich repertoire of infinite forms), and a
medium-size case system with both grammatical and adverbial functions,
and no gender distinction. They also show extensive contact with their
Germanic neighbours. Many grammatical structures are head-initial
constructions, as compared to the classical Uralic pattern, and the
tense system is compatible with the North Germanic one.

@@

Taking a closer look, there are still differences. In both North and
South Saami, negation is expressed with a negation verb which inflects
for person and number, but in 
South Saami, this verb is inflected for Tense as well. South Saami
has OV word order and lacks a copula in predicative constructions, 
whereas North Saami uses VO in simple sentences, and requires a copula.

@@

\section{Implementation}
%@fran
% [Pipeline image]

@@

\subsubsection{Lexical transfer}

Large syntactic differences also have consequences for the bidix, because there are many examples of sme-verb has go be translated into object+verb og adverb+verb in sma.
ex: nob `å trekke lodd' sme \textit{vuorbádit} sma \textit{vuerpiem giesedh}, 
nob `å presentere' sme \textit{ovdanbuktit} sma \textit{åvtese buektedh}. 
I mange tilfeller er grunnen til denne  forskjellen at man på sme har fått utviklet flere termer som matcher de norske teremene, mens man på sma
fremdeles må forklare innholdet.

>>>>

\usepackage{xltxtra}
\usepackage{fontspec}

@@

\usepackage[small,bf]{caption}

\setmainfont[BoldFont={Times New Roman}]{Times New Roman}
\setromanfont{Times New Roman}
%\setmonofont{FreeMono}

@@

\begin{abstract}
This paper describes the development of a rule-based
machine translation system from North Sámi to South
Sámi, and its evaluation in a setting where North Sámi
functions as pivot translation via manual translation
from Norwegian North Sámi, and thereafter MT to South Sámi. 
\end{abstract}

@@

%pivot translation references --babych paper
\cite{babych2007}

@@

As a Germanic language, Norwegian differs from the Saami languages
in numerous respects. It has definiteness as a morphological category,
no case, prepositions and verb / particle constructions, and a 
relatively strict word order, with V2 in main clauses. Norwegian
and the Saami languages also show Sprachbund phenomena, though.
Most notably, they have the same tense system. Whereas the Saami
languages also possess a number of infinite verb construction, the
Norwegian embedded clause pattern is in most cases a possible option
for the Saami languages as well.

The Saami languages constitute the westernmost branch of the Uralic
language family. They possess most of the classical Uralic
characteristica: A rich verbal morphology (3 numbers and persons, a
rich repertoire of infinite forms), and a medium-size case system with
both grammatical and adverbial functions, and no gender
distinction. They also show extensive contact with their Germanic
neighbours. Many grammatical structures are head-initial
constructions, as compared to the classical Uralic pattern, and the
tense system is compatible with the North Germanic one.

@@

Taking a closer look, there are still differences. In both North and
South Saami, negation is expressed with a negation verb which inflects
for person and number, but in South Saami, this verb is inflected for
Tense as well. South Saami has OV word order and lacks a copula in
predicative constructions, whereas North Saami uses VO in simple
sentences, and requires a copula.  The possession construction is
different from the Norwegian one, normally using copula rather than
possessor subject + a verb \textit{to have}. In North Saami, the
construction is 
\textit{locative possessor + copula + possessed in nominative}, 
whereas the possessor is in the genitive in South
Saami.

@@

\section{Implementation}
%@fran

@@

\subsubsection{Lexical transfer}
%@lene
Large syntactic differences also have consequences for the bidix, because there are many examples of sme-verb has go be translated into object+verb og adverb+verb in sma.
ex: nob `å trekke lodd' sme \textit{vuorbádit} sma \textit{vuerpiem giesedh}, 
nob `å presentere' sme \textit{ovdanbuktit} sma \textit{åvtese buektedh}. 
I mange tilfeller er grunnen til denne  forskjellen at man på sme har fått utviklet flere termer som matcher de norske teremene, mens man på sma
fremdeles må forklare innholdet.
